---
title: "RMS Titanic"
author: "Attila Sharipbek"
date: 09/09/2026
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% glimpse()


```

**Observations**:

- Class, Sex, Age, Survived, n

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- There are 2201 total passengers on board the titanic according to the df_titanic data.
- Are there any differences?
  - According to the Wikipedia article, there were 2208 passengers and crew on board the Titanic. This makes a difference in 7 souls from the df_titanic data that is provided.
- If yes, what might account for those differences?
  - The small difference in souls could be from different approaches to estimation of passengers. They were many lower class passengers that could of not been in their offical logs.
  - It also could be from different historical sources that were compiled at different times. Wikipedia is constantly updated, however the df_titanic data is built from the 1920 British Board of Trade Inquiry, and later compiled which means the data is old and could have been updated.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`

df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

- The data shows us that in raw numbers, more female passengers survived in the 1st, 2nd, and 3rd class. In crew more male souls survived then females. However, if you were to plot using percentages, in all 4 classes, a higher percent in females survived, due to the Titanic's prioritization of female and child passengers when evacuating.

# Deeper Lookprioritization

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge", color = "black")

```

\

**Observations**:

- The graph is unclear for what its protraying. It shows that 1st class had 100% of something but unclear for what. Especially after adding the argument (color = "black"color ) shows that the proportions are stacked.
- Is there anything *fishy* going on in your plot?
  - The plot stacked the proportion of age data on top of the proportion of who survived which makes it look like stacked data, with no distinction of which proportion is which. Using facet_grid would better portray this.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}


df_titanic %>% 
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_grid(Age ~ Survived)
```

**Observations**:

- These graphs show the raw number of female, male, child, and adult passengers that did survive and didn't survive. Graph the shows adults that didnt survive, shows that there were for more male passengers on board then female and that most of the female passengers were saved, except for the 3rd class, which shows that they still had class priority even for the women. Similarly, the graph the shows children that didnt survive, only children in class 3 didnt survive.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - 
    - With 4 different distinct graphs to show class, sex and age, for who survived and who didn't. You can actually see the different data rather then stacking it all into one graph. If used the df_prop, it would display the proportions individually in their specific graph.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
