---
title: "Gapminder"
author: "(Attila Sharipbek)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
summary(gapminder)
glimpse(gapminder)
```

**Observations**:

country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot()


gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

**Observations**:

- In 1952 Oceania had the highest median GDP per capita, around \$10,000. Europe came next at about \$5,100, then the Americas at about \$3,000, with Asia around \$1,200 and Africa around \$1,000 at the bottom. Oceania's box is basically a flat line because it only has two countries. Asia has the widest spread, plus one huge outlier above \$100,000. Most of the other continents have a few high outliers too, so a continent's median hides some big gaps between its countries.

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them
- One Asian country sits above \$100,000, so the y-axis has to stretch way up and every other box gets squished near zero, which makes the continents hard to compare. Putting the y-axis on a log scale with scale_y_log10 spread the boxes back out.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2

gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
 
  filter(gdpPercap > quantile(gdpPercap, 0.75) + 1.5 * IQR(gdpPercap)) %>%
  arrange(continent, desc(gdpPercap))
```

**Observations**:

- Identify the outlier countries from q2
  - South Africa, Gabon, Angola, United States, Canada, Venezuela, Kuwait, Bahrain, Switzerland

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

outliers <- c(
  "South Africa", "Gabon", "Angola", "United States", "Canada",
  "Venezuela", "Kuwait", "Bahrain", "Switzerland"
)

gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% outliers),
    mapping = aes(color = country),
    size = 2
  ) +
  scale_y_log10() +
  facet_wrap(~ year)
```

**Observations**:

- From 1952 to 2007 every continent's median GDP per capita went up, though not evenly. Europe's grew about five times while Africa's barely changed, so the gap between them got a lot wider. Kuwait is the only highlighted country that dropped, from about \$108,000 to \$47,000, and it doesn't stick out from everyone else the way it did in 1952. The US and Canada are still the top two in the Americas, and Gabon is now the richest country in Africa

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  
  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  facet_wrap(~ year)
```

- Did life expectancy gaps grow the way the GDP gaps did? Not really. Every continent's median went up, and Asia jumped the most, from about 45 to 72 years, almost catching the Americas. Africa is still lowest at about 53, and its box got about twice as wide, so some African countries improved a lot while others barely moved.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(year == year_max) %>%
  
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(mapping = aes(color = continent)) +
  scale_x_log10()
```

- Do people in richer countries live longer? In 2007 they mostly do: the points climb steadily as GDP per capita goes up. A few African countries break the pattern, though. They have GDP per capita over \$9,000 but life expectancy around 50.

```{r q5-task3}
gapminder %>%
  filter(continent == "Africa") %>%
  
  ggplot(aes(year, lifeExp)) +
  geom_point() +
  geom_point(
    data = . %>% filter(country %in% c("Zimbabwe", "Swaziland", "Botswana", "South Africa")),
    mapping = aes(color = country),
    size = 2
  )
```

- Did life expectancy in Africa rise steadily? Until about 1990, yes. After that the best countries kept improving but the middle stalled around 52, and the highlighted countries went backwards after peaking around 1987–1992. Zimbabwe fell from about 62 to 43 and Swaziland from about 59 to 40 by 2007. The lone point near 24 in 1992 is Rwanda, the lowest value in the whole dataset.
